---
title: "hw4_Adam_Banwell"
author: "Adam Banwell"
date: "10/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Commands

## Part one: Cleaning data
```{r}
library(dplyr)
library(ggplot2)

library(readxl)
defense <- read_excel('cyclonesFootball2020.xlsx', sheet='Defensive')
defense
str(defense)

offense <- read_excel('cyclonesFootball2020.xlsx', sheet='Offensive')
offense
str(offense)

biography <- read_excel('cyclonesFootball2020.xlsx', sheet='Biography')
biography
str(biography)

```
#1. player names and opponent names should be factors.
```{r}
defense$Name = factor(defense$Name)
defense$Opponent_Opponent = factor(defense$Opponent_Opponent)

offense$Name = factor(offense$Name)
offense$Opponent_Opponent = factor(offense$Opponent_Opponent)

biography$Name = factor(biography$Name)
```

#2. All offensive and defensive statistics should be numerical. The Weight column in biogrpahy should also be numeric.
```{r}
defClean = defense %>% select(-Name, -Opponent_Opponent) %>% mutate_if(is.character,as.numeric)
defClean = defense %>% select(Name, Opponent_Opponent) %>% mutate(defClean)
defClean

#Replace Pass cmp-att with two seperate columns
offClean = offense %>% mutate(Passing_CMP = substring(offense$"Passing_CMP-ATT", 1,2))
offClean = offClean %>% mutate(Passing_ATT = substring(offense$"Passing_CMP-ATT", 57))

offclean = offClean %>% select(-"Passing_CMP-ATT")

offClean2 = offclean %>% select(-Name, -Opponent_Opponent) %>% mutate_if(is.character,as.numeric)
offClean = offClean %>% select(Name, Opponent_Opponent) %>% mutate(offClean2)
offClean

bioClean = biography
bioClean$Weight = as.numeric(bioClean$Weight)
bioClean

```

#3. The Height column in biography should be numeric.
```{r}
feet = substring(bioClean$Height, 1, 1)
inches = substring(bioClean$Height, 3)

inches = as.numeric(inches)

feet2inches = as.numeric(feet) * 12

total = feet2inches+inches

bioClean$Height = total

```

#4. Add a column to both defense and offense indicating whether it is the first or the second game versus an opponent.
```{r}
defClean = defClean %>% group_by(Name)  %>% 
  mutate(game = case_when(
    duplicated(Opponent_Opponent)==TRUE ~ 2,
    duplicated(Opponent_Opponent)==FALSE ~ 1
    ))

offClean = offClean %>% group_by(Name)  %>% 
  mutate(game = case_when(
    duplicated(Opponent_Opponent)==TRUE ~ 2,
    duplicated(Opponent_Opponent)==FALSE ~ 1
    ))

str(defClean)
str(offClean)
str(bioClean)

```

##Part two: Tidying and exploring
#1. Reformat the offClean data frame into a tidy format using pivot_longer.
```{r}
library(tidyverse)
offClean = offClean %>%
  pivot_longer(!Name & !Opponent_Opponent, names_to = "stat", values_to = "values")
offClean

```

#2. Obtain the player-level offensive statistics. Namely, for each player and each type of statistics, obtain the sum of the statistics in all games.
```{r}
stats=offClean %>% drop_na()
stats=stats %>% group_by(Name, stat)
stats=stats %>% summarise(sum=sum(values))
stats=stats[!(stats$stat=="game"),]
stats
```

#3. Show the distributions of the player-level offensive statistics. The distribution of each offensive statsitic should be shown in a separate facet. Explain the patterns you see.
```{r}
stats=stats[!(stats$sum==0),]
stats
ggplot(stats, aes(x=Name, y=sum, fill = Name))+geom_bar(stat='identity')+facet_wrap(~stat, scales = "free")
```

#Did ISU have better offense against Oregon or Oklahoma? 
```{r}
oregon=offClean[(offClean$Opponent_Opponent=='Oregon'),]
oregon=oregon[(oregon$stat=='Receiving_YDS'),]
oklahoma=offClean[(offClean$Opponent_Opponent=='Oklahoma'),]
oklahoma=oklahoma[(oklahoma$stat=='Receiving_YDS'),]

oklahoma = rename(oklahoma, values2 = values)
vs = oregon %>% full_join(oklahoma,by="Name")
ggplot(data = vs, aes(x = values, y = values2)) + geom_point()

```
ISU had a better offense vs Oklahoma due to more yards in those games.

#4. Separate the Hometown column into two columns, namely, the city and the state where the player comes from. Print the first few rows of these two columns.
```{r}
bioClean
bioClean = bioClean %>% separate(Hometown, c('City', 'State'), sep=",")
bioClean
bioClean = bioClean %>% select(City, State)
print(bioClean, n=3)
```

#5. How many players are there from each state?
```{r}
bioClean
bioClean %>% group_by(State) %>% count()
```

#6. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team? Comment on both the defense and offense.
```{r}
offClean %>% group_by(Name) %>% filter(Name=="Purdy, Brock") %>% filter(stat=='Passing_CMP') %>%  summarise(Avg_Pass_CMP = mean(values, na.rm = TRUE))
```

Offense will directly do better when Purdy's completion status is high and offense does lower when he is low